Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-31 analysis/paper/paper.Rmd
fix typo and use American spelling

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeast Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeast Taiwan. We infer increasing craft specialization and changes in social organization. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
keywords: |
  Geometric Morphometrics; ceramic specialization; Iron age; culture contact; standardization
highlights: |
  These are the highlights. 
---

@@

The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society. We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene. Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001]. This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. That said, a more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994]. Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018]. Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015]. However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular. @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization. The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].

@@

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary. Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .

@@

The principal component (PC) scores were analysed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases. We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions. The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample. It is defined as the standard deviation divided by the mean in a ratio scale format:

@@

By standardizing standard deviation in each data set, the coefficients of variation (CV) values allows us to directly compare variation in samples measured with different units or means. CV values are robust to different sample sizes in the condition that the mean-standard deviation relationship is linear and positive since standard deviation is scaled to the mean [@Eerkens2001, pp.496]. We take this as our measurement of standardisation in vessel shape variables: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.

@@

To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period. Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

@@

# Normalise the shapes, from 
# https://www.repository.cam.ac.uk/bitstream/handle/1810/266423/Supplementary%20Code%20S1.pdf?sequence=20&isAllowed=y
jpgs_imported_coo <- 
  jpgs_imported %>% 
   Out(., fac = phases_isect)  %>% 
  # Centring, scaling and sampling pseudo-landmarks
   coo_center %>% 
   coo_scale %>% 
   coo_sample(., 1000) %>% 
  # Procrustes superimposition for pseudo-landmarks
   fgProcrustes %>% 
  # Normalisation of the starting points
   coo_slidedirection

@@

```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}
# what are the variences for each PC in each period?
# CVs on shape PCs for standardisation
len <- length(op.p$x[,1])
# plot PC1-3 distribution by phase
pc_plot <- tibble(pc = c(op.p$x[,1],
                         op.p$x[,2],
                         op.p$x[,3]),
                  phase = rep(op.p$fac$phase, 3),
                  pcn = str_glue('PC{sort(rep(1:3, len))}'))

@@

Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

@@

Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. If increased shape standardisation is a reliable indicator of craft specialisation, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000]. However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence. This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods. We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011]. Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who preferred northern Taiwan as their trading base. Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf. @Acabado2017]. In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and Chinese in the 19th century appears to have been more intense and direct. Historical records indicate that Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993]. This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.

We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence. Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese. In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardisation, and resistance through the overall continuity in vessel shape. Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized. This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.

@@

This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeast Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional. The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

>>>>

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. We infer increasing craft specialization and changes in social organization. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
keywords: |
  Geometric Morphometrics; ceramic specialization; Iron age; culture contact; standardization
highlights: |
  These are the highlights. 
---

@@

The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society. We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene. Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001]. This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. That said, a more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994]. Craft specialization may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018]. Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015]. However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular. @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization. The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].

@@

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary. Here we analyze pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardization of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases. The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .

@@

The principal component (PC) scores were analyzed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases. We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions. The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample. It is defined as the standard deviation divided by the mean in a ratio scale format:

@@

By standardizing standard deviation in each data set, the coefficients of variation (CV) values allows us to directly compare variation in samples measured with different units or means. CV values are robust to different sample sizes in the condition that the mean-standard deviation relationship is linear and positive since standard deviation is scaled to the mean [@Eerkens2001, pp.496]. We take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.

@@

To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period. Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialization because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

@@

# Normalise the shapes, from 
# https://www.repository.cam.ac.uk/bitstream/handle/1810/266423/Supplementary%20Code%20S1.pdf?sequence=20&isAllowed=y
jpgs_imported_coo <- 
  jpgs_imported %>% 
   Out(., fac = phases_isect)  %>% 
  # Centring, scaling and sampling pseudo-landmarks
   coo_center %>% 
   coo_scale %>% 
   coo_sample(., 1000) %>% 
  # Procrustes superimposition for pseudo-landmarks
   fgProcrustes %>% 
  # Normalization of the starting points
   coo_slidedirection

@@

```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}
# what are the variences for each PC in each period?
# CVs on shape PCs for standardization
len <- length(op.p$x[,1])
# plot PC1-3 distribution by phase
pc_plot <- tibble(pc = c(op.p$x[,1],
                         op.p$x[,2],
                         op.p$x[,3]),
                  phase = rep(op.p$fac$phase, 3),
                  pcn = str_glue('PC{sort(rep(1:3, len))}'))

@@

Our CV tests indicates that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

@@

Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000]. However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence. This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods. We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011]. Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who preferred northern Taiwan as their trading base. Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf. @Acabado2017]. In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and Chinese in the 19th century appears to have been more intense and direct. Historical records indicate that Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993]. This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.

We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence. Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese. In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape. Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized. This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.

@@

This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeastern Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional. The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
